White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2018-01-22 papers/wdmerger_II/paper.tex
Clean up formatting

White dwarfs are mainly composed of $\alpha$-chain particles, primarily ${}^4$He,
${}^{12}$C, ${}^{16}$O, ${}^{20}$Ne, and ${}^{24}$ Mg. Therefore the core of
any network appropriate for modeling nuclear burning in white dwarfs will be
these alpha chain nuclides, with the idea being that links up the $\alpha$-chain
will eventually get us to ${}^{56}$Ni, the nuclide responsible for the
energy output of Type Ia supernovae. In this paper we use the \aproxthirteen\
network\ \citep{timmes:1999,timmes:2000}. This includes all thirteen $\alpha$-chain
particles between ${}^4$He and ${}^{56}$Ni. This network was used by \citet{hawley:2012}
and \citet{raskin:2010}. \citet{loren-aguilar:2010} and \citet{garcia-senz:2013} used a
very similar network that additionally included ${}^{60}$Zn. This network has been
ported into a form that is consistent with the \amrex\ codes, in the freely available
\microphysics\ code repository
\footnote{\microphysics\ can be obtained at \url{https://github.com/starkiller-astro/Microphysics}.},
a collection of microphysical routines that are designed to be used in our
hydrodynamics codes. In our \microphysics\ repository we have also ported \aproxnineteen\
\citep{timmes:1999}, which was used in a few previous collision papers, and \aproxtwentyone.
The isotopes added by these networks relative to \aproxthirteen\ are not particularly relevant
to the problem of head-on white dwarf collisions, so we do not discuss results with these networks
here (though we have verified for this problem that these larger networks do not significantly change
the answer). We also make available \isoseven\ \citep{timmes:2000}, which skips the nuclides between
silicon and nickel, and was used by \citet{rosswog:2009}. \isoseven\ gives significantly different
quantitative results than \aproxthirteen\ for this problem, due to the inapplicable assumption of
an equilibrium abundance of the unused isotopes, but the results are still qualitatively similar.

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations. The sensitivity
of results to the value of these timestep limiters will be discussed in 
\autoref{sec:temporalresolution}. The factors $f_{be}$ and $f_{bX}$ can be set at runtime in \castro.

>>>>

White dwarfs are primarily composed of the $\alpha$-chain particles ${}^4$He,
${}^{12}$C, ${}^{16}$O, ${}^{20}$Ne, and ${}^{24}$Mg. Therefore the core of
any network appropriate for modeling nuclear burning in white dwarfs will be
these alpha chain nuclides, with the idea being that links up the $\alpha$-chain
will eventually get us to ${}^{56}$Ni, the nuclide responsible for the
energy output of Type Ia supernovae. In this paper we use the \aproxthirteen\
network\ \citep{timmes:1999,timmes:2000}. This includes all thirteen $\alpha$-chain
particles between ${}^4$He and ${}^{56}$Ni. This network was used by \citet{hawley:2012}
and \citet{raskin:2010}. \citet{loren-aguilar:2010} and \citet{garcia-senz:2013} used a
very similar network that additionally included ${}^{60}$Zn. This network has been
ported into a form that is consistent with the \amrex\ codes, in the freely available
\microphysics\ code repository\footnote{\microphysics\ can be obtained at
\url{https://github.com/starkiller-astro/Microphysics}.},
a collection of microphysical routines that are designed to be used in our
hydrodynamics codes. In our \microphysics\ repository we have also ported \aproxnineteen\
\citep{timmes:1999}, which was used in a few previous collision papers, and \aproxtwentyone.
The isotopes added by these networks relative to \aproxthirteen\ are not particularly relevant
to the problem of head-on white dwarf collisions, so we do not discuss results with these networks
here (though we have verified for this problem that these larger networks do not significantly change
the answer). We also make available \isoseven\ \citep{timmes:2000}, which skips the nuclides between
silicon and nickel, and was used by \citet{rosswog:2009}. \isoseven\ gives significantly different
quantitative results than \aproxthirteen\ for this problem, due to the inapplicable assumption of
an equilibrium abundance of the unused isotopes, but the results are still qualitatively similar.

@@

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations. The sensitivity
of results to the value of these timestep limiters will be discussed later in the
results. The factors $f_{be}$ and $f_{bX}$ can be set at runtime in \castro.
